Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-06-25 paper.qmd
Corrected some spelling mistakes

```{r}
#| label: best-fit-tsne
#| fig-cap: "Lifting the \\gD{} fitted model into \\pD{}. Two projections of the \\pD{} fitted model overlaying the data are shown in b, c. The fit is reasonbly tight with the data in one cluster (top one in b), but slightly less so in the other cluster probably because it is in $3\\text{-}D$. Notice also that, in the \\gD{} layout the two clusters have internal gaps which creates a model with some holes. This lacy pattern happens regardless of the hyper-parameter choice, but this doen't severely impact the \\pD{} model representation."
#| fig-width: 8
#| fig-height: 2

@@

When examining the \gD{} model representations it appears to be flatter or like a pancake with some methods, especially PACMAP, when the data is structure is higher than two dimensional. A simple example of this can be seen with data simulated to contain five \fD{} Gaussian clusters. Each cluster is essentially a ball in \fD{}, so there is no \gD{} representation, rather the model in each cluster should resemble a crumpled sheet of paper that fills out \fD{}.

@@

```{r}
#| label: fig-five-gau-projs
#| fig-height: 10
#| fig-width: 15
#| fig-pos: H
#| fig-cap: "NLDR's organise points in the \\gD{} layout in different ways, possibly misleadingly, illustrated using three layouts: (a) tSNE, (b) UMAP, (c) PaCMAP. The data has five Gaussian cluster in \\fD{}. The bottom row of plots shows a \\gD{} projection from a tour on \\fD{} revealing the differences generated by the layouts on the model fits.  We would expect the model fit to be like that in (a2) where it is distinctly is separate for each cluster but like a hairball in each. This would indicate the distinct clusters, each being fully \\fD{}. With (c2), the curiousity is that the model is a \\gD{} pancake shape in \\fD{}, indicating that there is some ordering of points done by PaCMAP, posisbly along some principal component axes."

@@

```{r}
#| label: fig-one-dens_clust-error
#| fig-width: 15
#| fig-height: 10
#| fig-pos: H
#| out-width: 100%
#| fig-cap: "Exploring the effect of density on the NLDR layout using a C-shaped structure with different density at each end. The top part of the triangular shape, shown with linked brushing, identify high model error corresponding to the sparse end of the structure (a1-a3). On the other hand, one of the dense corners of the structure shows low model error and is highlighted in the right part of the triangular shape (b1-b3). The tSNE layouts (a1, b1) are colored according to selected points. a2, a3 and b2, b3 present the results from linked plots that brushed the \\fD{} error points in \\gD{} and \\fD{}. It helps in identifying the high \\fD{} model errors that occur due to the sparse end."

@@

This is a benchmark single-cell data set collected on Human Peripheral Blood Mononuclear Cells (PBMC3k) as used in @pbmc2019. Single-cell data measures the gene expression of individual cells in a sample of tissue (see  for example, @haque2017). This type of data is used to obtain an understanding of cellular level behavior and heterogeneity in their activity. Clustering of single-cell data is used to identify groups of cells with similar expression profiles. NLDR often used to summarize the cluster structure. Usually, NLDR does not use the cluster labels to compute the layout, but uses color to represent the cluster labels when it is plotted.

>>>>

```{r}
#| label: best-fit-tsne
#| fig-cap: "Lifting the \\gD{} fitted model into \\pD{}. Two projections of the \\pD{} fitted model overlaying the data are shown in b, c. The fit is reasonably tight with the data in one cluster (top one in b), but slightly less so in the other cluster probably because it is in $3\\text{-}D$. Notice also that, in the \\gD{} layout the two clusters have internal gaps which creates a model with some holes. This lacy pattern happens regardless of the hyper-parameter choice, but this doen't severely impact the \\pD{} model representation."
#| fig-width: 8
#| fig-height: 2

@@

When examining the \gD{} model representations it appears to be flatter or like a pancake with some methods, especially PaCMAP, when the data is structure is higher than two dimensional. A simple example of this can be seen with data simulated to contain five \fD{} Gaussian clusters. Each cluster is essentially a ball in \fD{}, so there is no \gD{} representation, rather the model in each cluster should resemble a crumpled sheet of paper that fills out \fD{}.

@@

```{r}
#| label: fig-five-gau-projs
#| fig-height: 10
#| fig-width: 15
#| fig-pos: H
#| fig-cap: "NLDR's organise points in the \\gD{} layout in different ways, possibly misleadingly, illustrated using three layouts: (a) tSNE, (b) UMAP, (c) PaCMAP. The data has five Gaussian clusters in \\fD{}. The bottom row of plots shows a \\gD{} projection from a tour on \\fD{} revealing the differences generated by the layouts on the model fits.  We would expect the model fit to be like that in (a2) where it is distinctly is separate for each cluster but like a hairball in each. This would indicate the distinct clusters, each being fully \\fD{}. With (c2), the curiousity is that the model is a \\gD{} pancake shape in \\fD{}, indicating that there is some ordering of points done by PaCMAP, posisbly along some principal component axes."

@@

```{r}
#| label: fig-one-dens_clust-error
#| fig-width: 15
#| fig-height: 10
#| fig-pos: H
#| out-width: 100%
#| fig-cap: "Exploring the effect of density on the NLDR layout using a C-shaped structure with different density at each end. The top part of the triangular shape, shown with linked brushing, identify high model error corresponding to the sparse end of the structure (a1-a3). On the other hand, one of the dense corners of the structure shows low model error and is highlighted in the right part of the triangular shape (b1-b3). The tSNE layouts (a2, b2) are colored according to selected points. a2, a3 and b2, b3 present the results from linked plots that brushed the \\fD{} error points in \\gD{} and \\fD{}. It helps in identifying the high \\fD{} model errors that occur due to the sparse end."

@@

This is a benchmark single-cell data set collected on Human Peripheral Blood Mononuclear Cells (PBMC3k) as used in @pbmc2019. Single-cell data measures the gene expression of individual cells in a sample of tissue (see  for example, @haque2017). This type of data is used to obtain an understanding of cellular level behavior and heterogeneity in their activity. Clustering of single-cell data is used to identify groups of cells with similar expression profiles. NLDR is often used to summarize the cluster structure. Usually, NLDR does not use the cluster labels to compute the layout, but uses color to represent the cluster labels when it is plotted.
